---
title: "Coursera Practical Machine Learning: Course Project"
author: "Todd Himple"
date: "January 27, 2017"
output: html_document
---


##Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect 
a large amount of data about personal activity relatively inexpensively. These type of devices 
are part of the quantified self movement - a group of enthusiasts who take measurements about 
themselves regularly to improve their health, to find patterns in their behavior, or because 
they are tech geeks. One thing that people regularly do is quantify how much of a particular 
activity they do, but they rarely quantify how well they do it. 

In this project, my goal is evaulate  data from accelerometers on the belt, forearm, arm, 
and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly 
in 5 different ways:

* Class A: exactly according to the specification
* Class B: throwing the elbows to the front
* Class C: lifting the dumbbell only halfway
* Class D: lowering the dumbbell only halfway
* Class E: throwing the hips to the front

More about this study can be found here:  http://groupware.les.inf.puc-rio.br/har#literature#ixzz4WEA9mGLy


##R Packages:
Installing possible packages that may be necessary for my analysis
```{r Packages, include=TRUE}
library(caret)
library(rpart) 
library(rpart.plot) 
library(RColorBrewer)
library(rattle)
library(randomForest)
```


##Datasets
Loading datasets and subsetting potentially useful data for analysis
```{r data processing, include=TRUE}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
valUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

Training_Data <- read.csv(url(trainUrl), header=TRUE)
train_subset <- Training_Data[,c(8:11,37:49,60:68,84:86,102,113:124,140,151:160)]

dim(train_subset)
#[1] 19622    53
#head(train_subset)
```


##Partitioning data
Creating two separate datasets, one for training (75%) and one for testing (25%)
```{r partitioning, include=TRUE}
inTrain <- createDataPartition(train_subset$classe, p = .75, list = FALSE)
training_data <- train_subset[inTrain, ]
testing_data <- train_subset[-inTrain, ]

#dim(training_data)
#dim(testing_data)
```


##Model training
Training model using Random Forest due to it's high accuracy.  The model is build on a 
training set of 19,622 records and 53 variables. 
```{r training, include=TRUE}
set.seed(123) #for reproducability
random_forest <- randomForest(classe~.,data=training_data,ntree=100,importance=TRUE)
random_forest

##Plotting the error rate as a function of the number of trees
plot(random_forest)
```


The model shows 99.55% accuracy on the training data.  Additionally, the plot shows the error
rate approaching new zero at it reaches 100 trees


##Validation
Training model using Random Forest due to it's high accuracy.  The model is build on a 
training set of 19,622 records and 53 variables. 
```{r validation, include=TRUE}
test_predictions <- predict(random_forest, newdata=testing_data)
confusionMatrix(test_predictions,testing_data$classe)
```
The test dataset was then applied to the trained model, which returned a very high 
accuracy of 99.59%



